MSstatsPTM statistical relative quantification of post-translational modifications in bottom-up proteomic experiments

### 2022-03-16 main/MSstatsPTM_Main.tex
Suggestion for the results

@@

Despite the important implications of PTMs in biological functions, there is a lack of general framework to summarize the available quantitative information from LC-MS data, to perform statistical inference, and to draw conclusions to characterize the quantitative properties of PTM in a statistically rigorous manner. Many investigations performed differential abundance analysis of PTMs using two-sample t-test or its extensions. The approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another. Modifications of the t-test such as moderated t-test with Limma were also proposed \cite{Zhu}. While simple, the approach does not fully account for the sources of variations, and it is not directly applicable to experiments with complex designs, e.g., comparisons of multiple conditions, acquisition in multiple batches, etc. Additionally, while this approach can be applied to experiments targeting PTMs, there is not a self contained, straight forward implementation of the methods, making application challenging.

Isobar-PTM was developed for experiments with MS/MS quantitative strategies that employ isobaric labels such as tandem mass tags (TMT) and isobaric tag for relative and absolute quantification (iTRAQ)\cite{Breitwieser:2013}. Isobar-PTM expresses MS measurements with a linear model and performs adjustment with respect to protein abundance using the difference between log-ratio of modified peptides in two channels and log-ratio of protein level. The modeling framework, however, is not applicable for either label-free workflows or experiments with complex designs.

Further details on the specific equations and workflow of existing methods can be seen in Supplementary Sec. \ref{supp-sec:ttest}.

@@

>>>>

\todo{Would it be possible to re-organize this section? (1) Overview, including the data structures and the figure; (2) Existing methods: limma and t-test, including the description in the notation of the figure; also Isobar-PTM - can express in the same notation? Ok if not; (3) Existing methods: MSstats. Not applicable for PTM but useful background; (4) Proposed approach in the notation of the figure. I feel that the experimental design in Suppler is redundant. Could we fold it into the main text as well?} \todo{In this way, the entire Suppler Sec 2 will be merged into main \; the only methods in supplementary will be the table }

@@

Despite the importance of PTMs for biological function, there is a lack of general statistical analysis framework that summarizes the available quantitative information from LC-MS/MS experiments, performs statistical inference, and characterizes the quantitative properties of PTM in a statistically rigorous manner.

Many investigations performed differential abundance analysis of PTMs using two-sample t-test or its extensions.  The approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another \todo{add references that substantiate that claim}. Modifications of the t-test such as moderated t-test with Limma were also proposed \cite{Zhu}. While simple, these approaches do not fully account for the sources of variations, and are not directly applicable to experiments with complex designs, e.g., comparisons of multiple conditions, acquisition in multiple batches, etc. Additionally, while these approach can be applied to experiments studying PTMs, there is not a self contained, straight forward implementation of the methods, making application challenging.

More recently, Isobar-PTM was proposed for experiments with LC-MS/MS quantitative strategies that employ isobaric labels such as tandem mass tags (TMT) or isobaric tag for relative and absolute quantification (iTRAQ)\cite{Breitwieser:2013}. Isobar-PTM expresses MS measurements with a linear model and performs adjustment with respect to protein abundance using the difference between log-ratio of modified peptides in two channels and log-ratio of protein level. Unfortunately, this statistical modeling framework is not applicable to either label-free workflows to experiments with complex designs.

Mathematical details regarding these existing methods is in Supplementary Sec. \ref{supp-sec:ttest}. \todo{Does the section describe Isobar-PTM as well?}

@@

\todo{I do not think you need to repeat all the MassIVE.quant ids - just refer to the table in the beginning. Then add that in addition to the data the repository containers also include analysis scripts and results}
